---
title: "Reproducible Research Project I"
author: "Kishore"
date: "August 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
My code begins here:
```{r,echo=TRUE}
    #setting up current working directory
    #setwd("c:/Users/kishored/Dropbox/MyWork/Reproducible_Research/Week2/Project1/")

    #downloading the downloader package
    #install.packages("downloader")

    #loading the package to download the activity data
    library(downloader)

    if(!file.exists('./Data/activity.csv')){
        dataset_url <-  "https://www.coursera.org/learn/reproducible-research/peer/gYyPt/course-project-1"
        download(dataset_url, dest = "data.zip", mode = "wb")
        unzip("data.zip", exdir = "./Data")

    }
    
    # Load the necessary library packages:
    library(ggplot2)
    library(plyr)
    
    #loading the activity data to work on
    activityData <- read.csv('./Data/activity.csv')
```
We now process the data:
```{r, echo=TRUE}
    activityData$day <- weekdays(as.Date(activityData$date))
    activityData$DateTime <- as.POSIXct(activityData$date, format="%Y-%m-%d")

    ##pulling data without nas
    cleanData <- activityData[!is.na(activityData$steps),]

```
* Calculate the total number of steps taken per day
* Summarize total steps per date:
```{r, echo=TRUE}
    sumSteps <- aggregate(activityData$steps ~ activityData$date, FUN=sum)
    colnames(sumSteps)<- c("Date", "Steps")
```
We are going to create a histogram now for the total number of steps taken each day:
```{r,echo=TRUE}
hist(sumSteps$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
```

* What is the mean of total number of steps taken per day?
Computing the mean and median of the total number of steps taken per day:
```{r, echo=TRUE}
as.integer(mean(sumSteps$Steps))

as.integer(median(sumSteps$Steps))
```

* What is the average daily activity pattern?
* Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r,echo=TRUE}
 
    ##create average number of steps per interval
    intervalData <- ddply(cleanData, .(interval), summarize, Avg = mean(steps))

    ##Create line plot of average number of steps per interval
    p <- ggplot(intervalData, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
    p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")
```


* Which 5-minute interval, on average across all the days in the dataset, 
  contains the maximum number of steps?
* Which interval contains the maximum average number of steps?

```{r,echo=TRUE}
    maxSteps <- max(intervalData$Avg)
    intervalData[intervalData$Avg==maxSteps,1]
```

We are nor going to impute missing values:
```{r,echo=TRUE}
    # Calculate and report the total number of missing values in the dataset
    # (i.e. # the total number of rows with NAs)
    # Number of NAs in original data set
    nrow(activityData[is.na(activityData$steps),])
    ## [1] 2304
    
    # The total number of rows with steps = 'NA' is 2304.
    # Devise a strategy for filling in all of the missing values in the dataset. 
    # The strategy does not need to be sophisticated. For example, you could use 
    # the mean/median for that day, or the mean for that 5-minute interval, etc.
    # My strategy for filling in NAs will be to substitute the missing steps with the average 5-minute interval based on the day of the week.
    # Create the average number of steps per weekday and interval
    avgTable <- ddply(cleanData, .(interval, day), summarize, Avg = mean(steps))
    
    ## Create dataset with all NAs for substitution
    naData<- activityData[is.na(activityData$steps),]
    
    ## Merge NA data with average weekday interval for substitution
    newData<-merge(naData, avgTable, by=c("interval", "day"))
    
    # Create a new dataset that is equal to the original dataset but with the missing data filled in.
    ## Reorder the new substituded data in the same format as clean data set
    newData2<- newData[,c(6,4,1,2,5)]
    colnames(newData2)<- c("steps", "date", "interval", "day", "DateTime")

    ##Merge the NA averages and non NA data together
    mergeData <- rbind(cleanData, newData2)
    # Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
    ##Create sum of steps per date to compare with step 1
    sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum)
    colnames(sumTable2)<- c("Date", "Steps")

    ## Mean of Steps with NA data taken care of
    as.integer(mean(sumTable2$Steps))
    ## [1] 10821
    
    ## Median of Steps with NA data taken care of
    as.integer(median(sumTable2$Steps))
    ## [1] 11015
    
    ## Creating the histogram of total steps per day, categorized by data set to show impact
    hist(sumTable2$Steps, breaks=5, xlab="Steps", 
         main = "Total Steps per Day with NAs Fixed", col="Black")
    hist(sumTable2$Steps, breaks=5, xlab="Steps", 
         main = "Total Steps per Day with NAs Fixed", col="Grey", add=T)
```
